Keep your Laziness in Check

### 2018-01-04 paper/main.tex
working on the intro

\begin{itemize}
\item Laziness is a fundamental feature in Haskell \leo{WhyFPmatters?}
\item However, reasoning about programs with laziness (performance, behavior) is hard.
\item It is very easy to use a too-lazy or too-strict implementation that yields different results.
\item FRP quote here?
\item
\item Example: folds.
\item \lk{fold} to calculate sum breaks for large lists because of thunks
\item \lk{fold'} is good for some, but not for average because of whnf.
\item We need an implementation that is strict in the pair as well.
\item
\item Which leads to the natural question, how can we test the laziness behavior of functions?
\item Even worse, how can we even {\em specify} the laziness behavior?
\end{itemize}

Our technical contributions are as follows:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior of functions that is
{\em precise} \leo{define}. \leo{Also add drawbacks/partiality here? If not here, where in the intro?}
\item We propose a random testing approach to test such specifications.
\item We implement StrictCheck. \leo{In order for these last two points to be different, we need
something about the cool/awesome hacks in the implementation to point out here}
\item As a side contribution? We fix the function generaton of QuickCheck to produce lazy/partial functions instead
of fully strict ones all the time.
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.

>>>>

Laziness is ubiquitous in modern software engineering. The big-data framework
Spark uses laziness to drive computations from demand (\hzh{find references to
spark}), Apple's new programming language Swift includes lazy evaluation for
initializing expensive resources only when they are needed, and many functional
reactive programming frameworks for the modern web use laziness.

However, reasoning about the performance of programs with laziness is very
hard. It is very easy to accidentally make a program too lazy or too strict, and
this often makes the program behave differently from what the programmer had
intended.

Since laziness is a fundamental feature of the programming language Haskell, we
will explore how to observe and specify laziness using just the features
provided by Haskell, without custom compiler support.

\leo{
\begin{itemize}
\item Laziness is a fundamental feature in Haskell (quote WhyFPmatters?)
\item However, reasoning about programs with laziness (performance, behavior) is hard.
\item It is very easy to use a too-lazy or too-strict implementation that yields different results.
\item FRP quote here?
\end{itemize}
}

Consider the following implementation of the \lk{sum} function:
\leo{I created an environment for this so that we can tweak them all as needed}
\begin{inlinecode}
sum = foldl (+) 0
\end{inlinecode}

Although this seemingly innocent implementation of sum will produce correct
results on small lists of numbers, it will quickly run out of memory on larger
lists. The culprit here is the strictness behavior of \lk{foldl}---it does not
force the accumulated result during the fold. Calling this \lk{sum} on a list
will produce a thunk that is linear in the size of of the input list.

We can fix this space leak for \lk{sum} using the strict version \lk{foldl'},
which forces the accumulator. However, even this is not enough in
general. Consider a single-traversal computation of an average where we keep
both a running sum and a running length in an accumulator.

\begin{inlinecode}
avg = uncurry (/) . foldl (\(s,n) x -> (s + x, n + 1)) (0,0)
\end{inlinecode}

In this case, both \lk{foldl} and \lk{foldl'} construct thunks linear in the
size of the input list, since forcing a pair into weak head normal form does not
force either component.

How can we test and catch these laziness bugs before they manifest themselves in
production? Is it possible to specify the laziness behavior using a precise
language just as we would for functional correctness?

\leo{We probably need a paragraph that explains what we want to do here,
not directly a jump into the contributions}

In this paper, we present solutions to these questions. We will demonstrate how
to observe laziness from within Haskell itself, and how to write down a
specification for laziness also using Haskell.

Our technical contributions are as follows:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior of functions that is
{\em precise} \leo{define}. \leo{Also add drawbacks/partiality here? If not here, where in the intro?}
\item We propose a random testing approach to test such specifications.
\item We demonstrate non-trivial generic programming techniques to apply StrictCheck to almost all data types. \leo{In order for these last two points to be different, we need
something about the cool/awesome hacks in the implementation to point out here}
\item As a side contribution? We fix the function generaton of QuickCheck to produce lazy/partial functions instead
of fully strict ones all the time.
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.
